---
output: github_document
---

<!-- 
README.md is generated from README.Rmd. Please edit that file
rmarkdown::render("README.Rmd") 
-->


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

<!-- badges: start -->
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/solartime)](http://cran.r-project.org/package=solartime)
[![R-CMD-check](https://github.com/bgctw/solartime/workflows/R-CMD-check/badge.svg)](https://github.com/bgctw/solartime/actions)
<!-- badges: end -->

## Overview

The `solartime` package provides utilities to work with solar time, 
i.e. where noon is exactly when sun culminates.
It provides functions to compute

- difference between local time zone and solar time
- sun position
- sunrise, sunset, daylength, and daytime flags

## Installation

```{r, eval = FALSE}
# From CRAN (in future)
#install.packages("solartime")

# Or the the development version from GitHub:
# install.packages("devtools")
devtools::install_github("bgctw/solartime")
```

## Usage

The city of Jena is located west of the timezone's meridian.
Hence, the sun culminates around 13 minutes later.

The time difference shifts during the year because of 
[earth orbit's eccentricity](https://en.wikipedia.org/wiki/Orbital_eccentricity).

```{r eccentricityPlot}
require(solartime, quietly = TRUE)
latDeg <- 50.93; longDeg <- 11.57
doy <- 1:366
locaDiffMin <- 60 * computeSolarToLocalTimeDifference(
  longDeg, timeZone = 1L, doy = doy)
plot(locaDiffMin  ~ doi, ylab = "time difference solar - local time (min)")
abline(h = localDiff); abline(h = 0, lty = "dotted")
```

See the [package vignettes](https://github.com/bgctw/solartime/tree/master/vignettes) (*.md) for further examples.

