---
output: github_document
---

<!-- 
README.md is generated from README.Rmd. Please edit that file
rmarkdown::render("README.Rmd") 
-->


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/solartime)](http://cran.r-project.org/package=solartime)
[![Travis-CI Build Status](https://travis-ci.org/bgctw/solartime.svg?branch=master)](https://travis-ci.org/bgctw/solartime)


## Overview

The `solartime` package provides utilities to work with solar time, 
i.e. where noon is exactly when sun culminates.
It provides functions to compute

- difference between local time zone and solar time
- sun position
- sunrise, sunset, daylength, and daytime flags

## Installation

```{r, eval = FALSE}
# From CRAN (in future)
#install.packages("solartime")

# Or the the development version from GitHub:
# install.packages("devtools")
devtools::install_github("bgctw/solartime")
```

## Usage

```{r}
require(solartime, quietly = TRUE)
```

### Correcting for difference of solar time to local time

The city of Jena is located west of the timezone's meridian.
Hence, the sun culminates around 13 minutes later.

```{r}
# Jena lat, long: 50.927222, 11.586111
latDeg <- 50.93
longDeg <- 11.57
# its local time zone is one hour ahead of UTC
timeZone <- +1
(localDiff <- computeSolarToLocalTimeDifference(longDeg, 1L)*60)
```
The time difference shifts during the year because of 
[earth orbit's eccentricity](https://en.wikipedia.org/wiki/Orbital_eccentricity).

```{r}
doi <- 1:366
locaDiffDoi <- computeSolarToLocalTimeDifference(longDeg, 1L, doi)*60
plot(locaDiffDoi  ~ doi, ylab = "time difference solar - local time (min)")
abline(h = localDiff); abline(h = 0, lty = "dotted")
```

During most days, the sun culmintes after noon, but during a few days 
in autumn the sun culminates earlier than noon of the lcoal time zone.


### Computing sun position

Using function `computeSunPosition`.

```{r}
times <- seq( 
  ISOdate(2018, 11, 21, 0, tz = "Etc/GMT-1"), by = "2 hour", length.out = 13)
ans <- computeSunPosition(times, latDeg = latDeg, longDeg = longDeg)
cbind(data.frame(timestamp = times), as.data.frame(ans))
```

The return value is a data.frame with polar coordinates of the sun in radian
on the [horizontal coordinatesystem](https://en.wikipedia.org/wiki/Horizontal_coordinate_system): 

- elevation: angle above the horizon, 
- azimuth: angle around the horizon, measured from true north increasing 
  eastward, and 
- [declination](https://en.wikipedia.org/wiki/Declination): angle between
  celestial equator and the observers, fundamental plane, i.e. the plane that
  contains the horizon
- fractional hour after midnight
    
In the example the azimuth increases from slightly more than zero at midnight to 
2$\pi$ 
at the following midnight. Elevation increases from negative values to 0 at 
sunset, 
maximum at noon declining to zero at sunset. Declination in early winter 
decreases to
more negative values until it reaches the negative of the earth axial tilt on 
December solistice.


### Sunrise and sunset

Sunrise and sunset are computed in fractional hours after midnight. 
Neglecting the difference between solar time and local time introduces a bias.
Daylength is not biased by neglecting solar time correction.

```{r}
today <- as.POSIXlt(ISOdate(2018,3,1,0, tz = "Etc/GMT-1"))
(sunrise <- computeSunriseHour(today, latDeg = latDeg
                               , isCorrectSolartime = FALSE))
(sunrise <- computeSunriseHour(today, latDeg = latDeg, longDeg = 11.586))
(sunset <- computeSunsetHour(today, latDeg = latDeg, longDeg = 11.586))
(daylength <- computeDayLength(today, latDeg = latDeg))
```

Sunrise is set to 12 for polar nights and 0 for polar days.
Similarly, sunset is set to 12 for polar nights and to 0 for polar days.
```{r}
doy <- 1:366
sunrise <- computeSunriseHourDoy( doy, latDeg = +80, isCorrectSolartime = FALSE)
sunset <- computeSunsetHourDoy( doy, latDeg = +80, isCorrectSolartime = FALSE)
par(mfrow = c(1,2)); plot(sunrise ~ doy ); plot(sunset ~ doy )
```

### Further Utilities

Functions `computeIsDayByHour` and `computeIsDayByLocation` 
quickly classify daytime and nighttime records of a dataset.

```{r}
dateSeq <- seq( 
  as.POSIXct("2017-03-20", tz = "Etc/GMT-1")
  , as.POSIXct("2017-03-21", tz = "Etc/GMT-1")
  , by = "30 min")
isDay <- computeIsDayByLocation(dateSeq, latDeg = 50.93, longDeg = 11.59)
```

```{r echo=FALSE}
plot( isDay ~ dateSeq )
yday <- as.POSIXlt(dateSeq[1])$yday + 1L
sunrise <- computeSunriseHourDoy(
  yday, latDeg = 50.93, longDeg = 11.59, timeZone = 1)
sunset <- computeSunsetHourDoy(
  yday, latDeg = 50.93, longDeg = 11.59, timeZone = 1)
abline( v = trunc(dateSeq[1], units = "days") + c(sunrise,sunset)*3600L )
```

Function `getHoursAheadOfUTC` provides the integer timeZone argument 
of a timestamp, required many other functions of the package.
Similarly, `getFractionalHours` provides fractional hours after midnight of
a timestamp.

```{r}
#around daylight saving time step in Central European time
tzBack <- Sys.getenv('TZ')
Sys.setenv(TZ = 'CET')
#Sys.setenv(TZ = 'UTC')
t1 <- as.POSIXct("2018-10-28 01:30")
(dateSeq <- seq(t1, by = "30 min", length.out = 6))
(hoursAhead <- getHoursAheadOfUTC(dateSeq))
(facHours <- getFractionalHours(dateSeq))
Sys.setenv(TZ = tzBack)
```

<!-- 
See the [package vignettes](https://github.com/bgctw/solartime/tree/master/vignettes) (*.md) for further examples.
-->

