---
output: 
  rmarkdown::html_vignette:
    keep_md: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteIndexEntry{overview}
  %\usepackage[UTF-8]{inputenc}
---

```{r, include = FALSE}
# do not execute on CRAN: 
# https://stackoverflow.com/questions/28961431/computationally-heavy-r-vignettes
#is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
#             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
#knitr::opts_chunk$set(eval = !is_check)
```

```{r setup, include = FALSE}
#rmarkdown::render("vignettes/overview.Rmd")
knitr::opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(
    fig.align = "center"
    #, fig.width = 3.27, fig.height = 2.5, dev.args = list(pointsize = 10)
    #,cache = TRUE
    , fig.width = 4.3, fig.height = 3.2, dev.args = list(pointsize = 10)
    #, fig.width = 6.3, fig.height = 6.2, dev.args = list(pointsize = 10)
    # works with html but causes problems with latex
    #,out.extra = 'style = "display:block; margin: auto"' 
    )
knitr::knit_hooks$set(spar = function(before, options, envir) {
    if (before) {
        par(las = 1 )                   #also y axis labels horizontal
        par(mar = c(1.8,2.3,0,0) + 0.3 )  #margins
        par(tck = 0.02 )                          #axe-tick length inside plots             
        par(mgp = c(1.1,0.2,0) )  #positioning of axis title, axis labels, axis
     }
})
```


## Overview

The `solartime` package provides utilities to work with solar time, 
i.e. where noon is exactly when sun culminates.
It provides functions to compute

- difference between local time zone and solar time
- sun position
- sunrise, sunset, daylength, and daytime flags

## Usage

```{r}
require(solartime)
```

### Difference between solar time and local time

The city of Jena is located west of the timezone's meridian.
Hence, the sun culminates around 13 minutes later.

```{r}
latDeg <- 50.93; longDeg <- 11.57
(localDiff <- computeSolarToLocalTimeDifference(longDeg, 1L)*60)
```
The time difference shifts during the year because of 
[earth orbit's eccentricity](https://en.wikipedia.org/wiki/Orbital_eccentricity).

```{r}
doy <- 1:366
locaDiffDoi <- computeSolarToLocalTimeDifference(longDeg, 1L, doy)*60
plot(locaDiffDoi  ~ doy, ylab = "time difference solar - local time (min)")
abline(h = localDiff); abline(h = 0, lty = "dotted")
```

During most days, the sun culmintes after noon, but during a few days 
in autumn the sun culminates earlier than noon of the lcoal time zone.


### Computing sun position

Using function `computeSunPosition`.

```{r}
times <- seq( 
  ISOdate(2018, 11, 21, 0, tz = "Etc/GMT-1"), by = "2 hour", length.out = 13)
ans <- computeSunPosition(times, latDeg = latDeg, longDeg = longDeg)
cbind(data.frame(timestamp = times), as.data.frame(ans))
```

The return value is a data.frame with polar coordinates of the sun in radian
on the [horizontal coordinatesystem](https://en.wikipedia.org/wiki/Horizontal_coordinate_system): 

- elevation: angle above the horizon, 
- azimuth: angle around the horizon, measured from true north increasing 
  eastward, and 
- [declination](https://en.wikipedia.org/wiki/Declination): angle between
  celestial equator and the observers, fundamental plane, i.e. the plane that
  contains the horizon
- fractional hour after midnight
    
In the example the azimuth increases from slightly more than zero at midnight to 
2$\pi$ 
at the following midnight. Elevation increases from negative values to 0 at 
sunset, 
maximum at noon declining to zero at sunset. Declination in early winter 
decreases to
more negative values until it reaches the negative of the earth axial tilt on 
December solistice.


### Sunrise and sunset

Sunrise and sunset are computed in fractional hours after midnight. 
Neglecting the difference between solar time and local time introduces a bias.
Daylength is not biased by neglecting solar time correction.

```{r}
today <- as.POSIXlt(ISOdate(2018,3,1,0, tz = "Etc/GMT-1"))
(sunrise <- computeSunriseHour(today, latDeg = latDeg
                               , isCorrectSolartime = FALSE))
(sunrise <- computeSunriseHour(today, latDeg = latDeg, longDeg = 11.586))
(sunset <- computeSunsetHour(today, latDeg = latDeg, longDeg = 11.586))
(daylength <- computeDayLength(today, latDeg = latDeg))
```

Sunrise is set to 12 for polar nights and 0 for polar days.
Similarly, sunset is set to 12 for polar nights and to 0 for polar days.
```{r}
doy <- 1:366
sunrise <- computeSunriseHourDoy( doy, latDeg = +80, isCorrectSolartime = FALSE)
sunset <- computeSunsetHourDoy( doy, latDeg = +80, isCorrectSolartime = FALSE)
par(mfrow = c(1,2)); plot(sunrise ~ doy ); plot(sunset ~ doy )
```

### Further Utilities

Functions `computeIsDayByHour` and `computeIsDayByLocation` 
quickly classify daytime and nighttime records of a dataset.

```{r}
dateSeq <- seq( 
  as.POSIXct("2017-03-20", tz = "Etc/GMT-1")
  , as.POSIXct("2017-03-21", tz = "Etc/GMT-1")
  , by = "30 min")
isDay <- computeIsDayByLocation(dateSeq, latDeg = 50.93, longDeg = 11.59)
```

```{r echo=FALSE}
plot( isDay ~ dateSeq )
sunrise <- computeSunriseHour(dateSeq[1], latDeg = 50.93, longDeg = 11.59)
sunset <- computeSunsetHour(dateSeq[1], latDeg = 50.93, longDeg = 11.59)
abline( v = trunc(dateSeq[1], units = "days") + c(sunrise,sunset)*3600L )
```

Function `getHoursAheadOfUTC` provides the integer timeZone argument 
of a timestamp, required other functions of the package.
Similarly, `getFractionalHours` provides fractional hours after midnight of
a timestamp.

```{r}
#around daylight saving time step in Central European time
tzBack <- Sys.getenv('TZ'); Sys.setenv(TZ = 'CET') 
  t1 <- as.POSIXct("2018-10-28 01:30")
  (dateSeq <- seq(t1, by = "30 min", length.out = 6))
  (hoursAhead <- getHoursAheadOfUTC(dateSeq))
  (facHours <- getFractionalHours(dateSeq))
Sys.setenv(TZ = tzBack)
```

<!-- 
See the [package vignettes](https://github.com/bgctw/solartime/tree/master/vignettes) (*.md) for further examples.
-->

